Processing genome scale tabular data with wormtable

### 2013-08-17 docs/paper/paper.tex
Added some comments to paper.tex
Incorporated Dan's comments.

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
\paragraph*{Background:} 
Biological research generates large amounts of tabular data.
These tables are produced by 
programs and exported to plain text files, which are often very large. 
Researchers must process these files 
line-by-line, parsing values encoded as text into native types so that 
calculations may be done. This is a time consuming process.
Furthermore, there is no simple means of indexing these files so that 
rows containing particular values can be quickly found. 
\paragraph*{Results:} 
We introduce a new 
data format and software library called wormtable, which provides efficient 
access to tabular data in Python. Wormtable stores data in a compact
binary format, provides random access to rows, and enables sophisticated
indexing on columns within these tables. Files written in existing 
formats can be easily converted to wormtable format, and we 
provide conversion utilities for the VCF and GTF format.

@@

\emph{The Background section should be written in a
way that is accessible to researchers without specialist knowledge in
that area and must clearly state - and, if helpful, illustrate - the
background to the research and its aims. It should clearly described the
relevant context and the specific issue which the software described is
intended to address.}

Despite the ever increasing volumes of data being processed in bioinformatics,
the methods used are almost entirely based on plain text files. Data is 
usually encoded in lines of text, with each row consisting of a series of tab-delimited 
values.
These files are easy to view and interpret and can be processed on 
any platform with the minimum of library dependencies. Using compression,
text files can be quite compact, and Tabix~\cite{li11} provides  
fast retrieval of rows from arbitrary locations in the table.

It is not sufficient, however, to simply store and retrieve data. To use data
it must be \emph{processed}. This is the major flaw in 
using text files as a data format: before we can perform calculations, we 
must first parse the encoded information into native 
machine values. This is an expensive process, and 
compression (if it is used) adds another substantial overhead. The result 
of this is that simple calculations over a large dataset may take 
many hours to complete.

Another problem with tables stored as text files is that it is difficult 
to index the information within particular columns. 
This means that many operations on a table require a complete scan 
through the file. As a result, the only viable means of 
working with a subset of the data
is to create another file consisting of the subset of interest. 
This is inflexible and error prone, and multiplies already significant  
storage requirements.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Implementation 
%%
\section*{Implementation}
\emph{This should include a description of the overall architecture of the software implementation, 
along with details of any critical issues and how they were addressed.
}

% what is wormtable?
Wormtable (write-once read-many table) 
is a new data format and software library designed specifically
to deal with the challenges of data processing in bioinformatics.
It provides a portable, compact, read-only store for tabular
data of essentially unlimited size. Tables can
be either written directly via the library, or converted
from existing formats via command line tools. 
Wormtable provides a simple Python API to access rows
in the table, facilitating efficient and convenient access to 
data. The core data manipulation and storage facilities 
are written in C for efficiency.

% Wormtable's binary row format 
Data is stored in rows using a  
compact binary format designed specifically for biological applications.
Each row consists of a fixed number of columns, and each column 
holds either a fixed or variable number of elements of a fixed type and size.
Wormtable currently supports signed and unsigned integers, 
floating point and character data. To store data as 
compactly as possible, integer sizes range from one to eight bytes,
and real numbers are stored as IEEE half, single, and double 
precision floating point values. 
Fixed and variable length character strings 
are also supported.

@@

% Indexes
Besides random access to rows, wormtable also provides indexes
over arbitrary combinations of columns. An index is a database in 
which the keys are the values from the columns in question concatenated
together, and the values are pointers to the original row. Indexes make
many operations much more efficient, since we can directly seek to the 
rows we are interested in without performing a full table scan. They also 
provide a very efficient means of calculating a histogram for a given 
set of columns. Wormtable also supports binned indexes, in which a range of 
values is mapped to a single index key. This is very useful for floating 
point columns, where it may not be necessary to distinguish between
values that differ by small amounts.

@@

\emph{The Results and Discussion may be combined into a single section or
presented separately. They may also be broken into subsections with
short, informative headings. In any case what should be described is the
functionality of the software together with data on how its performance
and functionality compare with and improve on functionally similar
existing software. There should then be a discussion of the intended
use of the software, and the benefits that are envisioned together,
if possible, with an outline for the planned future development of
new features.}

\subsection*{API}
The principle goal of wormtable is to provide efficient access to data 
using an easy to learn interface. 
The \texttt{Table} class provides the main interface,
and implements the standard Python sequence protocols. Efficient 
iteration over rows is provided by the \texttt{cursor} method, 
which takes as an argument the list of columns to read.
Only the values for the columns 
of interest are then retrieved, leading to considerable
time savings. The \texttt{cursor} function takes two additional 
arguments, \texttt{start} and \texttt{stop}, which specify 
rows of interest. This allows us to efficiently seek to an 
arbitary location in the table and to read a given number 
of rows sequentially from this point.

@@

The API is straightforward, but it is flexible, powerful and 
extensible. This simplicity ensures that 
programmer of all experience levels  
can take full advantage of the powerful data 
processing that wormtable provides. In the following subsections 
we illustrate the performance advantages of wormtable via some 
examples. These are not intended to be definitive benchmarks
and are simply concrete examples to demonstrate the type of improvements 
that can be expected by using wormtable over existing methods.

@@

Values are stored in wormtable in a portable binary format, so that no parsing is
required when reading in rows. To illustrate this advantage, we
wrote a script to count the number of transitions and transversions
in the dataset using wormtable and PyVCF~\cite{pyvcf}
a Python VCF parser.  In this example we proceed 
row-by-row, examining the REF and ALT columns and
counting the transitions and transversions we encounter. 
Using PyVCF this required approximately 126 minutes, whereas 
the wormtable version required
57 seconds. To compare against methods that are known to be extremely
efficient, we repeated the same 
example using the Unix tools \texttt{cut}, \texttt{grep} 
and \texttt{awk},  which required 80 seconds. Both the
Unix pipeline and wormtable were limited by the sequential read
bandwidth of the hard drive, and would therefore be much 
faster using modern solid state storage.
It should be noted, however, that although Unix 
pipelines are efficient, it is a difficult and error prone method 
of processing data when more complex calculations are required.

@@

To compare the seek performance of wormtable with Tabix, we compressed
the VCF file mentioned in the previous subsection using \texttt{bgzip}
and indexed it with \texttt{tabix}. Since the VCF file was not 
strictly in order of genome position (in violation of the VCF format),
we first needed to sort the VCF file, which required 5m13s. Afterwards,
compression required 11m8s, and indexing with \texttt{tabix} required
1m12s. The size of the resulting compressed file was 2.9GB and 
the size of the index file was 106KB.
To duplicate the functionality of Tabix on VCF data, we simply need to 
create an index on the CHROM and POS columns using \texttt{wtadmin add}. 
This required approximately
2 minutes and the resulting index consumed 115MB of space.

@@

When using a cold cache (i.e., no pages of the files in question are 
present in the operating system's cache) the elapsed
time for Tabix was 103 seconds with a processor time of 65
seconds. For wormtable, the elapsed time was 181 seconds 
with a processor time of 5 seconds. Thus, the time required 
to perform this test is dominated by waiting for I/O in both 
cases. Since the wormtable file is considerably larger than the 
compressed file used by Tabix, more and larger seeks were required 
to bring the required pages into memory. However, once the 
pages were in memory, Tabix needed to do much more work to 
decompress then and make them usable, as shown by the 
difference in processor times.

@@

The penalty for random seeks is exceedingly high in this 
example, however, since we have used a hard disk for storage.
Repeating the cold cache example on a laptop equipped with 
solid state storage, the elapsed time for Tabix was 76 seconds
with a CPU time of 72 seconds, whereas the elapsed time for 
wormtable was 31 seconds with a CPU time of 6.5 seconds. Thus, using 
solid state storage negates the advantage of smaller files, since 
seeks are much less expensive.  
 With the rapidly reducing cost and 
increasing capacity of solid state storage, it seems likely 
that it will soon be routinely used for all but the largest 
of bulk storage. Wormtable takes much better advantage of
modern hardware and large cache sizes, allowing the full bandwidth 
and random seek performance to be utilised. Compression based 
techniques will always be limited by the CPU time required to 
decompress data.

@@

The library supports efficient access to any data stored in 
wormtable format, and currently provides conversion utilitites 
for the VCF and GTF formats.
The most important aspect of future development 
is to develop tools to convert other tabular formats such as 
PSL, GFF and BED to wormtable format. Such tools are 
not difficult to develop, since all that is required is a parser
for the format in question written in Python.

@@

%%%%%%%%%%%%%%%%%%%%%%
\section*{Conclusions}
\emph{This should state clearly the main conclusions of the article and give a clear 
explanation of the importance and relevance of the software.}

The volumes of data being produced and processed in bioinformatics are constantly 
increasing, but the tools available to end users to 
process this data are still mostly based on parsing plain text. This 
approach is very inefficient, and leads to several undesirable outcomes.
Firstly, and most obviously, a researcher's productivity is inevitably
reduced by having to wait for several hours to see the result of a 
simple calculation. There are also some less obvious consequences:
code quality is reduced, since testing over the entire dataset is 
infeasible, and it is much less likely that the effects of 
changing arbitrary quality thresholds (for example) will be 
systematically examined.

We have concentrated on the VCF format here since it is a widely 
used and popular format. After converting the VCF format to 
wormtable using \texttt{vcf2wt} a user can process the 
data very efficiently using Python. Accessing data from 
rows in wormtable is many times faster than is possible 
by parsing rows encoded as text.
This is also a very convenient 
way to access VCF data, since individual columns are already 
parsed and all that is needed is the name of the columns of 
interest. For accessing regions of the genome efficiently
the user simply needs to create an index on the CHROM and POS columns
using the  \texttt{wtadmin add} command, and can then efficiently
seek to arbitrary locations. In the worst case, when using a 
slow hard disk this is somewhat slower than Tabix. In the best
case, when frequently accessing the same regions or using 
fast solid state storage, wormtable is many times faster than 
Tabix. Moreover, wormtable is much more general than Tabix, 
since it can index any column or combination of columns.
Similarly, wormtable is not limited to VCF, but can store
any form of fixed tabular data. We also provide a conversion 
tool for the GTF format, and several others are planned.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Authors contributions}
    Text for this section \ldots

>>>>

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
\paragraph*{Background:} 
Biological research generates large amounts of tabular data.
These tables are produced by 
programs and exported to plain text files, which are often very large. 
Researchers must process these files 
line-by-line, parsing values encoded as text into native types so that 
calculations may be done. This is an error prone and time consuming process.
Furthermore, there is no simple means of indexing these files so that 
rows containing particular values can be quickly found. 
\paragraph*{Results:} 
We introduce a new 
data format and software library called wormtable, which provides efficient 
access to tabular data in Python. Wormtable stores data in a compact
binary format, provides random access to rows, and enables sophisticated
indexing on columns within these tables. Files written in existing 
formats can be easily converted to wormtable format, and we 
provide conversion utilities for the VCF and GTF format.

@@

%\emph{The Background section should be written in a
%way that is accessible to researchers without specialist knowledge in
%that area and must clearly state - and, if helpful, illustrate - the
%background to the research and its aims. It should clearly described the
%relevant context and the specific issue which the software described is
%intended to address.}
%
Despite the ever increasing volumes of data being processed in bioinformatics,
the methods used are almost entirely based on plain text files. Data is 
usually encoded in lines of text, with each row consisting of a series of tab-delimited 
values.
These files are easy to view and interpret and can be processed on 
any platform with the minimum of library dependencies. Using compression,
text files can be quite compact, and specialised indexing methods 
are available to quickly retrieve rows intersecting with a 
given genomic interval~\cite{li11}.

It is not sufficient, however, to simply store and retrieve data. To use data
it must be \emph{processed}. This is the major flaw in 
using text files as a data format: before we can perform calculations, we 
must first parse the encoded information into native 
machine values. This is an expensive process, and 
compression (if it is used) adds another substantial overhead. 
As a result,  simple calculations over a large dataset may take 
many hours to complete.

Another problem with tables stored as text files is that it is difficult 
to index the information within particular columns. 
This means that many operations on a table require a complete scan 
through the file. The only viable means of 
working with a subset of the data, therefore, 
is to create another file consisting of the subset of interest. 
This is inflexible and error prone, and multiplies already significant  
storage requirements.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Implementation 
%%
\section*{Implementation}
%\emph{This should include a description of the overall architecture of the software implementation, 
%along with details of any critical issues and how they were addressed.
%}

% what is wormtable?
Wormtable (write-once read-many table) 
is a new data format and software library designed specifically
to deal with the challenges of data processing in bioinformatics.
It provides a portable, compact, read-only store for tabular
data of essentially unlimited size. Tables can
be either written directly using the library, or converted
from existing formats via command line tools. 
Wormtable provides a simple Python API to access rows
in the table, facilitating efficient and convenient access to 
data. The core data manipulation and storage facilities 
are written in C for efficiency.

% Wormtable's binary row format 
Data is stored in rows using a compact binary format.
Each row consists of a fixed number of columns, and each column 
holds either a fixed or variable number of elements of a fixed type and size.
Each column is allocated a fixed number of bytes within the `fixed region' of 
a row. For fixed
length columns, the encoded elements are stored entirely within the
fixed region; for variable length columns, the encoded elements are 
stored in the `variable region' of the row, and the location and number 
of elements stored in the fixed region. The type, size and number 
of elements of 
each column are specified in the table schema, which is defined 
in XML as part of the table's metadata.

Wormtable currently supports signed and unsigned integers, 
floating point and character data. Integer sizes range 
from one to eight bytes, and real numbers are stored as 
IEEE half, single, and double precision~\cite{ieee08}
floating point values.  
Fixed and variable length character strings 
are also supported. Using this flexible type system, columns can 
be assigned the most appropriate type and size for the range of
values to be represented.

@@

% Indexes
Besides random access to rows, wormtable also provides indexes
over arbitrary combinations of columns. An index is a 
Berkeley DB database in 
which the keys are the elements from the columns in question concatenated
together, and the values are pointers to the original row. Indexes make
many operations much more efficient, since we can directly seek to the 
rows we are interested in without performing a full table scan. They also 
provide a very efficient means of calculating a histogram for a given 
set of columns. Wormtable also supports binned indexes, in which a range of 
values is mapped to a single index key. This is very useful for floating 
point columns, where it may not be necessary to distinguish between
values that differ by small amounts.

@@

%\emph{The Results and Discussion may be combined into a single section or
%presented separately. They may also be broken into subsections with
%short, informative headings. In any case what should be described is the
%functionality of the software together with data on how its performance
%and functionality compare with and improve on functionally similar
%existing software. There should then be a discussion of the intended
%use of the software, and the benefits that are envisioned together,
%if possible, with an outline for the planned future development of
%new features.}
%
\subsection*{API}
The principle goal of wormtable is to provide efficient access to data 
using an easy to learn interface. 
The \texttt{Table} class provides the main interface,
and implements the standard Python sequence protocols. Efficient 
iteration over rows is provided by the \texttt{cursor} method, 
which takes as an argument the list of columns to read.
Only the values for the columns 
of interest are then retrieved, leading to considerable
time savings. The \texttt{cursor} function takes two additional 
arguments, \texttt{start} and \texttt{stop}, which specify 
rows of interest. This allows us to efficiently seek to an 
arbitary location in the table and to read a given number 
of rows sequentially from this point.

@@

The API is straightforward, but it is flexible, powerful and 
extensible. This simplicity ensures that 
programmers of all experience levels  
can take full advantage of the powerful data 
processing facilities that wormtable provides. In the following subsections 
we illustrate the performance advantages of wormtable via some 
examples. These are not intended to be definitive benchmarks
and are simply concrete examples to demonstrate the type of improvements 
that can be expected by using wormtable over existing methods.

@@

Values are stored in wormtable in a portable binary format, so that no parsing is
required when reading in rows. To illustrate this advantage, we
wrote a script to count the number of transitions and transversions
in the dataset using wormtable and PyVCF~\cite{pyvcf},
a Python VCF parser.  In this example we proceed 
row-by-row, examining the REF and ALT columns and
counting the transitions and transversions we encounter. 
Using PyVCF this required approximately 126 minutes, whereas 
the wormtable version required
57 seconds. To compare against methods that are known to be extremely
efficient, we repeated the same 
example using the Unix tools \texttt{cut}, \texttt{grep} 
and \texttt{awk},  which required 80 seconds. Both the
Unix pipeline and wormtable were limited by the sequential read
bandwidth of the hard drive, and would therefore be much 
faster using modern solid state storage.
It should be noted, however, that although Unix 
pipelines are efficient, it is a difficult and error prone method 
of processing data when more complex calculations are required.

@@

To compare the seek performance of wormtable with Tabix, we compressed
the VCF file mentioned in the previous subsection using \texttt{bgzip}
(which required 11m8s) and indexed it with \texttt{tabix} (1m12s). 
The size of the resulting compressed file was 2.9GB and 
the size of the index file was 106KB.
To duplicate the functionality of Tabix on VCF data, we simply need to 
create an index on the CHROM and POS columns using \texttt{wtadmin add}. 
This required approximately
2 minutes and the resulting index consumed 115MB of space.

@@

When using a cold cache (i.e., no pages of the files in question are 
present in the operating system's cache) the elapsed
time for Tabix was 103 seconds with a processor time of 65
seconds. For wormtable, the elapsed time was 181 seconds 
with a processor time of 5 seconds. Thus, the time required 
to perform this test is dominated by waiting for I/O in both 
cases. Since the wormtable file is considerably larger than the 
compressed file used by Tabix, more and larger hard drive
seeks were required 
to bring the required pages into memory (solid state 
storage does not suffer from this problem). Once the 
pages were in memory, however, Tabix needed to do much more work to 
decompress then and make them usable, as shown by the 
difference in processor times.

@@

The test used here is also highly synthetic, and unlikely to be 
indicative of most real-world applications. The majority of 
workloads have strong locality of reference~\cite{d05}, 
and such large and extreme seeks across genomic regions 
are unlikely to occur. In this case, wormtable 
is  much faster than Tabix, since there
is no CPU overhead of decompression. Furthermore, since 
Tabix returns rows of text, the problem of parsing rows
must still be solved. This is expensive (as illustrated 
in the previous subsection), and creates extra code 
complexity. Wormtable, by providing a simple API to 
access both rows and columns, gives a 
unified interface for accessing data that is 
both straightforward to use and highly efficient.

@@

The library supports efficient access to any data stored in 
wormtable format, and currently provides conversion utilitites 
for the VCF and GTF formats.
The most important aspect of future development 
is to develop tools to convert other tabular formats such as 
PSL, GFF, SAM and BED to wormtable format. 
Such tools are not difficult to develop, since all that is required is a parser
for the format in question written in Python.

@@

%%%%%%%%%%%%%%%%%%%%%%
\section*{Conclusions}
%\emph{This should state clearly the main conclusions of the article and give a clear 
%explanation of the importance and relevance of the software.}
%
The volume of data being produced in bioinformatics is constantly 
increasing, but the tools available to end users to 
process data are still mostly based on parsing plain text. This 
approach is very inefficient, and leads to several undesirable outcomes.
Firstly, and most obviously, a researcher's productivity is inevitably
reduced by being forced to wait for several hours for the result of a 
simple calculation. Without flexible indexing, working with 
a subset of a data file usually requires the creation of 
another file consisting of the subset in question, requiring 
extra storage and maintenance.
There are also some less obvious consequences:
code quality is reduced, since testing over the entire dataset is 
infeasible, and it is much less likely that the effects of 
changing arbitrary quality thresholds (for example) will be 
systematically examined.

The classical approach to solving problems of this type 
is to use a relational database, which provide sophisticated 
data management techniques. Relational databases are unsuitable 
for storing static datasets, however, as they are complex to 
use and incur many unnecessary overheads.
Wormtable provides the most important features of database
technologies (packed binary storage of values; random access 
to rows; general purpose indexing) without their complexities 
and overheads. Wormtable is also more flexible than 
relational databases, supporting, for example, columns 
containing a variable number of integers. Finally, wormtable 
is far more adaptable than a relational database.
Any widely used database system today is a hugely complex 
piece of software, and adding necessary features would be 
very difficult. Adding new features to wormtable, on the 
other hand, is straightforward since it is far simpler and 
does not need to be compatible with the relational model
and decades worth of existing software.

The most important aspect of wormtable is its efficiency
and ease of use for end users, and we illustrated these
points using some examples of VCF data.
After converting a file in VCF format to 
wormtable using \texttt{vcf2wt} a user can process the 
data very efficiently using Python. Accessing data from 
rows in wormtable is many times faster than is possible 
by parsing rows encoded as text.
This is also a very convenient 
way to access VCF data, since individual columns are already 
parsed and all that is needed is the name of the column of 
interest. To access regions of the genome efficiently
the user simply needs to create an index on the CHROM and POS columns
using the  \texttt{wtadmin add} command, and can then efficiently
seek to arbitrary locations. Wormtable is not 
limited to VCF, but can store
any form of fixed tabular data. We also provide a conversion 
tool for the GTF format, and several others are planned.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Authors contributions}
JK conceived the project, developed wormtable source code and associated
scripts and drafted the manuscript. RN wrote the tutorial and contributed to
the example scripts and the manuscript. DH wrote the example scripts and vcf-utils,
and contributed to vcf2wt, the tutorial and the manuscript.
All authors read and approved the final manuscript.
